Computing volume via non-linear rectangular infinitesimals

### 2019-08-14 Cav2D/article-template.tex
made some progress with prismatoid section

\noindent
Moreover, assume that $\mathbf{c}(t)$ is a translational curve with respect to $f(x,y)$ on $R$. Let us first partition $R$ into $n\times m$ subrectangles $R_{ij} = [x_{i},x_{i+1}] \times [y_{j},y_{j+1}]$ of equal width $\Delta x^1 = \Delta x_i^1 = x_{i+1}-x_i\ll (b-a)$ and length $\Delta y^1= \Delta y_j^1 = y_{j+1}-y_j\ll(d-c)$.
Furthermore, let the transformation $T$ map the $xy$--coordinate $(x_i^1,y_i^1)\in R$ (the lower left vertice of $R_{ij}$) at which $\mathbf{c}(t) + \left <x_i^1,y_i^1,0 \right >$ intersects the 
$xy$-plane to the $xy$--coordinate at which $\mathbf{c}(t) + \left <x_i^1,y_j^1,0 \right >$ intersects $f(x,y)$. Note, the domain of $T$ is $R$. Moreover, let us denote the range 
of $T$ with $S$. As the paartition of $R$ can be made arbitrarily fine, $T:R\rightarrow S$ actually maps every point $(x_i^1,y_i^1)\in R$ to a point $(x_i^2,y_i^2)\in S$. In general, the transformation $T$ is defined as follows:
\begin{equation}
x_i^2 = h(x_i^1,y_i^1)~\textrm{and}~y_i^2 = s(x_i^1,y_i^1).
\end{equation}

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. 
As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
Let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\ref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_i^i$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_i^i$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_i^2)$ represents the height of a non-linear rectangular column or prism $\mathcal{K}_{ij}$ whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_i^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \lim_{n,m\rightarrow \infty} \sum_{j=1}^n\sum_{i=1}^n f(x_{ij}^2,y_{ij}^2) \Delta x_{ij}^1\Delta y_{ij}^1\\
&=&  \lim_{n,m\rightarrow \infty} \sum_{j=1}^n\sum_{i=1}^n f(h(x_{ij}^1,y_{ij}^1),s(x_{ij}^1,y_{ij}^1)) \Delta x_{ij}^1\Delta y_{ij}^1\\
&=& \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy.
\end{eqnarray}

\noindent
The only thing that now remains is to find expressions for $h$ and $s$. Note that, for every coordinate pair $(x_i^1,y_i^1)\in R$ and $(x_i^2,y_i^2)\in S$ there exists a unique $t\in \mathbb{R}$ such that 
\begin{equation}
\label{eq:vector_2}
\left< x_i^2, y_i^2, f(x_i^2,y_i^2) \right > = \left < x_i^1, y_i^1, 0 \right > + \left <c_x(t),c_y(t),c_z(t)\right >  
\end{equation}
The above equation can be re-expressed as:
\begin{equation}
c_z(t) = f(x_i^1 + c_x(t),y_i^1 + c_y(t)). 
\end{equation}
We can now solve for $t$ in terms of $x_i^1$ and $y_i^1$. If we substitute this solution into Equation~\ref{eq:vector_2} we obtain:
\begin{equation}
x_i^2 = h(x_i^1,y_i^1) = x_i^1 + c_x\circ t(x_i^1,y_i^1) 
\end{equation}
and 
\begin{equation}
y_i^2 = s(x_i^1,y_i^1) = y_i^1 + c_y\circ t(x_i^1,y_i^1). 
\end{equation}

\section{Prismatoid}
Let $\mathcal{S}$ be the solid which is bounded by the planes $f(x,y)=-x-y+8$, $2x$, $2(x-4)$, $y$, $y-4$ and the $xy$-plane. Solid $\mathcal{S}$ is depicted 
in Figure~\ref{fig:prismatoid_solid}, Figure~\ref{fig:prismatoid} and Figure~\ref{fig:prismatoid_real}. Solid $\mathcal{S}$ is a prismatoid, since all of its 
vertices lie in two parallel planes. Furthermore, all of its lateral faces are trapezoids or triangles (if viewed from the side). Also note the following: the base of $\mathcal{S}$ is the region 
 $R = [0,4]\times [0,4]$ and that the line of intersection between $2x$ and $y$ is $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$. The region $R$ and the line $\mathbf{c}(t)$ are
also depicted and labelled in Figure~\ref{fig:prismatoid_solid}.\\

\noindent
As we alluded to in the previous section we can calculate $V_{\mathcal{S}}$ using the double Cavalieri integral:
\begin{equation}
\iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We can, however, only evaluate the above integral if the transformation $T$ is known (see Figure ). We can find $T$ using Equation

Let $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ be the line of intersection between the planes $2x$ and $y$. Furthermore, note that the line 
$\mathbf{c}(t)$ passes through the origin. The line $\mathbf{c}(t)$ is depicted and labelled in Figure~\ref{fig:prismatoid_solid}. Let 
\begin{equation}
R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber
\end{equation}
and
\begin{equation}
S = \{(x,y)\in\mathbb{R}^2|-\sfrac{1}{3}y+\sfrac{8}{3}\leq x\leq -\sfrac{1}{3}y+\sfrac{16}{3}, \sfrac{1}{2} x + 4\leq y \leq \sfrac{1}{2} x + 6 \}. \nonumber
\end{equation}
The region $S$ and $R$ are depicted in Figure~\ref{fig:prismatoid_solid} and Figure~\ref{fig:prismatoid_regions}. Let us now subdivide $R$ into $n\times m$
subrectangles $R_{ij} = [x_{i},x_{i+1}] \times [y_{j},y_{j+1}]$ of equal width $\Delta x^1 = \Delta x_i^1 = x_{i+1}-x_i<<(b-a)$ and length $\Delta y^1= \Delta y_j^1 = y_{j+1}-y_j<<(d-c)$. 
Furthermore, let the transformation $T$ map the $xy$--coordinate $(x_i^1,y_i^1)\in R$ (the lower left vertice of $R_{ij}$) at which $\mathbf{c}(t) + \left <x_i^1,y_i^1,0 \right >$ intersects the 
$xy$-plane to the $xy$--coordinate $(x_i^2,y_j^2)\in S$ at which $\mathbf{c}(t) + \left <x_i^1,y_j^1,0 \right >$ intersects $f(x,y)$ (Figure~\ref{fig:prismatoid_solid} and Figure~\ref{fig:prismatoid_regions}). 
%The transformation $T$ is defined by the following equations:
%\begin{equation}
%x_i^2 = h(x_i^1,y_i^1)
%\end{equation}
%and
%\begin{equation}
%y_i^2 = s(x_i^1,y_i^1)
%\end{equation}

>>>>

\noindent
Moreover, assume that $\mathbf{c}(t)$ is a translational curve with respect to $f(x,y)$ on $R$. Let us first partition $R$ into $n\times m$ subrectangles $R_{ij} = [x_{i},x_{i+1}] \times [y_{j},y_{j+1}]$ of equal width $\Delta x^1 = \Delta x_i^1 = x_{i+1}-x_i\ll (b-a)$ and length $\Delta y^1= \Delta y_j^1 = y_{j+1}-y_j\ll(d-c)$.
Furthermore, let the transformation $T$ map the $xy$--coordinate $(x_i^1,y_j^1)\in R$ (the lower left vertice of $R_{ij}$) at which $\mathbf{c}(t) + \left <x_i^1,y_j^1,0 \right >$ intersects the 
$xy$-plane to the $xy$--coordinate at which $\mathbf{c}(t) + \left <x_i^1,y_j^1,0 \right >$ intersects $f(x,y)$. Note, the domain of $T$ is $R$. Moreover, let us denote the range 
of $T$ with $S$. As the paartition of $R$ can be made arbitrarily fine, $T:R\rightarrow S$ actually maps every point $(x_i^1,y_j^1)\in R$ to a point $(x_i^2,y_j^2)\in S$. In general, the transformation $T$ is defined as follows:
\begin{equation}
x_i^2 = h(x_i^1,y_j^1)~\textrm{and}~y_j^2 = s(x_i^1,y_j^1).
\end{equation}

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. 
As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
Let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\ref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_j^1$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_j^1$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_j^2)$ represents the height of a non-linear rectangular column or prism $\mathcal{K}_{ij}$ whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_j^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \lim_{n,m\rightarrow \infty} \sum_{j=1}^n\sum_{i=1}^n f(x_i^2,y_j^2) \Delta x_i^1\Delta y_j^1\\
&=&  \lim_{n,m\rightarrow \infty} \sum_{j=1}^n\sum_{i=1}^n f(h(x_i^1,y_j^1),s(x_i^1,y_j^1)) \Delta x_i^1\Delta y_j^1\\
&=& \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy.
\end{eqnarray}

\noindent
The only thing that now remains is to find expressions for $h$ and $s$. Note that, for every coordinate pair $(x_i^1,y_j^1)\in R$ and $(x_i^2,y_j^2)\in S$ there exists a unique $t\in \mathbb{R}$ such that 
\begin{equation}
\label{eq:vector_2}
\left< x_i^2, y_j^2, f(x_i^2,y_j^2) \right > = \left < x_i^1, y_j^1, 0 \right > + \left <c_x(t),c_y(t),c_z(t)\right >  
\end{equation}
The above equation can be re-expressed as:
\begin{equation}
\label{eq:c_z}
c_z(t) = f(x_i^1 + c_x(t),y_j^1 + c_y(t)). 
\end{equation}
We can now solve for $t$ in terms of $x_i^1$ and $y_j^1$. If we substitute this solution into Equation~\ref{eq:vector_2} we obtain:
\begin{equation}
\label{eq:h_2d}
x_i^2 = h(x_i^1,y_j^1) = x_i^1 + c_x\circ t(x_i^1,y_j^1) 
\end{equation}
and 
\begin{equation}
\label{eq:s_2d}
y_j^2 = s(x_i^1,y_j^1) = y_j^1 + c_y\circ t(x_i^1,y_j^1). 
\end{equation}

\section{Prismatoid}
Let $\mathcal{S}$ be the solid which is bounded by the planes $f(x,y)=-x-y+8$, $2x$, $2(x-4)$, $y$, $y-4$ and the $xy$-plane. Solid $\mathcal{S}$ is depicted 
in Figure~\ref{fig:prismatoid_solid}, Figure~\ref{fig:prismatoid} and Figure~\ref{fig:prismatoid_real}. Solid $\mathcal{S}$ is a prismatoid, since all of its 
vertices lie in two parallel planes. Furthermore, all of its lateral faces are trapezoids or triangles (if viewed from the top). Also note the following: the base of $\mathcal{S}$ is the region 
 $R = [0,4]\times [0,4]$ (if viewed from the front) and that the line of intersection between $2x$ and $y$ is $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$. The region $R$ and the line $\mathbf{c}(t)$ are
also depicted and labelled in Figure~\ref{fig:prismatoid_solid}.\\

\noindent
As we alluded to in the previous section we can calculate $V_{\mathcal{S}}$ using the double Cavalieri integral:
\begin{equation}
\label{eq:cav_ex1}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We can, however, only evaluate the above integral if the transformation $T$ is known (see Figure~\ref{fig:prismatoid_regions}). We can find $T$ using Equation~\eqref{eq:h_2d} and 
Equation~\eqref{eq:s_2d}. To evaluate Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} we first need to express $t$ in terms of $x_i^1$ and $y_j^1$. If we substitute $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t(x_i^1,y_j^1) = \frac{x_i^1 + y_j^1 + 8}{5} 
\end{equation}
Substituting $t(x_i^1,y_j^1)$ into Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} results in:
\begin{equation}
x_{i}^2 = h(x_{i}^1,y_{j}^1) = \frac{4}{5} x_{i}^1 - \frac{1}{5} y_{j}^1 + \frac{8}{5}  
\end{equation}
and
\begin{equation}
y_{j}^2 = r(x_{i}^1,y_{j}^1) = \frac{3}{5} x_{i}^1 - \frac{2}{5} y_{j}^1 + \frac{16}{5}.  
\end{equation}
We can now evaluate Equation~\eqref{eq:cav_ex1}:
\begin{eqnarray}
V_{\mathcal{S}}=\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^4\int_0^4 f(h(x,y),r(x,y))~dydx\\
&=& \int_0^4 \int_0^4 -\frac{7}{5}x + \frac{3}{5} y + \frac{16}{5}~dydx\\
&=& 25.6
\end{eqnarray}

We can easily verify that the $V_{\mathcal{S}}$ we computed is correct, since S is a prismatoid. According to the prismatoid volume 
equation $V_{\mathcal{S}}=25.6$. The details of this computation is described in the caption of Figure~\ref{fig:prismatoid}. This result verifies 
the fact that

Let $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ be the line of intersection between the planes $2x$ and $y$. Furthermore, note that the line 
$\mathbf{c}(t)$ passes through the origin. The line $\mathbf{c}(t)$ is depicted and labelled in Figure~\ref{fig:prismatoid_solid}. Let 
\begin{equation}
R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber
\end{equation}
and
\begin{equation}
S = \{(x,y)\in\mathbb{R}^2|-\sfrac{1}{3}y+\sfrac{8}{3}\leq x\leq -\sfrac{1}{3}y+\sfrac{16}{3}, \sfrac{1}{2} x + 4\leq y \leq \sfrac{1}{2} x + 6 \}. \nonumber
\end{equation}
The region $S$ and $R$ are depicted in Figure~\ref{fig:prismatoid_solid} and Figure~\ref{fig:prismatoid_regions}. Let us now subdivide $R$ into $n\times m$
subrectangles $R_{ij} = [x_{i},x_{i+1}] \times [y_{j},y_{j+1}]$ of equal width $\Delta x^1 = \Delta x_i^1 = x_{i+1}-x_i<<(b-a)$ and length $\Delta y^1= \Delta y_j^1 = y_{j+1}-y_j<<(d-c)$. 
Furthermore, let the transformation $T$ map the $xy$--coordinate $(x_i^1,y_j^1)\in R$ (the lower left vertice of $R_{ij}$) at which $\mathbf{c}(t) + \left <x_i^1,y_j^1,0 \right >$ intersects the 
$xy$-plane to the $xy$--coordinate $(x_i^2,y_j^2)\in S$ at which $\mathbf{c}(t) + \left <x_i^1,y_j^1,0 \right >$ intersects $f(x,y)$ (Figure~\ref{fig:prismatoid_solid} and Figure~\ref{fig:prismatoid_regions}). 
%The transformation $T$ is defined by the following equations:
%\begin{equation}
%x_i^2 = h(x_i^1,y_j^1)
%\end{equation}
%and
%\begin{equation}
%y_j^2 = s(x_i^1,y_j^1)
%\end{equation}
